---
title: "Assignment 1"
author: "Reuven Herzog"
date: "9/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(tidyverse)
library(tidycensus)
library(ggplot2)
```
Selecting data for census tracts in the 5 most urbanized counties in northern New Jersey: Bergen, Passaic, Essex, Hudson, and Union.
Selecting for urbanicity, majority race, percent renters, average household size, total population, and median age.
```{r Select Data, echo=FALSE}
counties <- c("Bergen","Passaic","Essex","Hudson","Union")
vars <- c(tot_hh = "H001001",
           tot_rural_hh = "H002005",
           tot_rent_hh = "H004004",
           avg_hh_size = "H012001",
           tot_pop = "P003001",
           nh_white_pop = "P005003",
           nh_black_pop = "P005004",
           nh_asian_pop = "P005006",
           hispanic_pop = "P005010",
           med_age = "P013001")
 
NNJ_tracts <- get_decennial(geography = "tract", variables = vars,
                            state = "NJ", county = counties,
                            output = "wide") %>%
     mutate(rural = (tot_rural_hh == tot_hh) ,
            pct_rental = tot_rent_hh / tot_hh,
            majority_race_eth = case_when(
                nh_white_pop / tot_pop > 0.5 ~ "nh_white",
                nh_black_pop / tot_pop > 0.5 ~ "nh_black",
                nh_asian_pop / tot_pop > 0.5 ~ "nh_asian",
                hispanic_pop / tot_pop > 0.5 ~ "hispanic",
                (nh_white_pop + nh_black_pop + nh_asian_pop + hispanic_pop) /
                    tot_pop < 0.5 ~ "other",
                TRUE ~ "None"),
            county = str_split_fixed(str_split_fixed(NAME,",",3)[,2]," ",3)[,2]) %>%
     filter(tot_pop > 0, tot_hh > 0) %>%
     select(GEOID, NAME, county, rural, majority_race_eth, 
            pct_rental, avg_hh_size, tot_pop, med_age) 
```

## 1) Histogram of the percent of renters per tract:
```{r histogram pct_rent, echo=FALSE}
ggplot(NNJ_tracts,aes(pct_rental))+
  geom_histogram(binwidth=0.033)
```

## 2) 2-d scatter plot
```{r scatter renter vs total, echo=FALSE}
ggplot(NNJ_tracts, 
       aes(x = tot_pop,
           y = pct_rental)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Tract population") +
  scale_y_continuous(name = "Share of Rental Households per Tract")
```

(sometimes the correlation just isn't there)


Maybe household size with rental percentage? And let's make it a log plot, just to test my comfort with that feature.
```{r scatter renter vs household size, echo=FALSE}
ggplot(NNJ_tracts, 
       aes(x = avg_hh_size,
           y = pct_rental)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5, method="lm") +
  scale_x_continuous(name = "Average Household Size") +
  scale_y_continuous(name = "Share of Rental Households per Tract", trans = "log")
```

```{r}
ggplot(NNJ_tracts, aes(x=pct_rental, y=med_age, color=majority_race_eth)) +
  geom_point() +
  stat_smooth(aes(group=1),method="lm", color="black") +
  scale_x_continuous(name = "Share of Rental Households") +
  scale_y_continuous(name = "Median Age") +
  scale_color_discrete(name = "Majority Race/Ethnicity",
                       labels = c("Hispanic", "Non-Hispanic Asian",
                       "Non-Hispanic Black", "Non-Hispanic White",
                       "No Majority"))
```

And for a twist - let's now sort by county instead of race:
```{r}
ggplot(NNJ_tracts, aes(x=pct_rental, y=med_age, color=county)) +
  geom_point() +
  stat_smooth(aes(group=1),method="lm", color="black") +
  scale_x_continuous(name = "Share of Rental Households") +
  scale_y_continuous(name = "Median Age") +
  scale_color_brewer(name = "County", palette = "Set1")
```


```{r size color madness, echo = FALSE}
ggplot(NNJ_tracts,aes(x = 0, y=0, color = med_age,size = tot_pop)) +
  geom_jitter() +
  scale_color_gradient(low = "green", high = "red") + 
  scale_x_continuous(name = "", labels = NULL, breaks = NULL) + 
  scale_y_continuous(name = "", labels = NULL, breaks = NULL)
```

OK this not only makes no sense, it's not really readable...

```{r aggregate, echo = FALSE}
v_01 <- aggregate(x = NNJ_tracts$tot_pop,
                  by = list(NNJ_tracts$majority_race_eth), FUN = sum) %>%
  rename(majority_race_eth = Group.1, tot_pop = x)
v_02 <- aggregate(x = NNJ_tracts$pct_rental,
                  by = list(NNJ_tracts$majority_race_eth), FUN = mean) %>%
  rename(majority_race_eth = Group.1, pct_rental = x)
summary <- merge(v_01,v_02,by = "majority_race_eth")
ggplot(summary, aes(x = majority_race_eth, y = pct_rental, fill = tot_pop)) +
         geom_col(width = 1) + coord_polar() +
  scale_x_discrete(name = "Majority race/ethnicity", labels = c("Hispanic", "Asian", "Black", "White", "No majority")) +
  scale_y_continuous(name = "Share of Rental Households") +
  scale_fill_continuous(name = "Total Population")
```
# Finally, a plot that is useful
```{r renters by county, echo=FALSE}
county_violin <- ggplot(NNJ_tracts,aes(x = reorder(county, -1*pct_rental), y = pct_rental)) +
  geom_violin(aes(fill=county, color=county),show.legend=FALSE) +
  scale_x_discrete(name=NULL) +
  scale_y_continuous(name="Share of Rental Households")
county_violin
```

This plot is very instructional. Both Hudson and Essex counties are highly urbanized, focused on Jersey City and Newark, respectively. It therefore makes sense that their share of rental households is above 50%. But Hudson is small enough to hardly include any non-urbanized tracts, whereas Essex includes less urban and suburban areas. Particularly interesting is Passaic County, whose southern section is composed of dense cities like Paterson and Passaic, with a relatively poor population - and thus many renters. However, its northern component is closer to rural, and it has many suburban single-family areas as well. Bergen similarly has two concentrations - one in mostly rental neighborhoods (though at a lower concentration than even Passaic), and a much larger bump of tracts with hardly any rental households.

##Violin plot with a box plot overlay
```{r}
county_violin + geom_boxplot(width = 0.075)
```

This diagram shows how boxplots often can hide the whole story, and in this case, I don't think it adds much beyond the violin plot.


#County rental rates - adding in ethnicity
```{r}
ggplot(NNJ_tracts,aes(x = reorder(county, -1*pct_rental), y = pct_rental, color = majority_race_eth)) +
  geom_jitter(width = 0.3) +
  scale_x_discrete(name=NULL) +
  scale_y_continuous(name="Share of Rental Households") +
  scale_color_discrete(name = "Majority Race/Ethnicity",
                       labels = c("Hispanic", "Non-Hispanic Asian",
                       "Non-Hispanic Black", "Non-Hispanic White",
                       "No Majority"))
```

This graph is particularly interesting. For one thing, I didn't realize that the Black community was so concentrated in Essex county, to the exclusion of other counties. (For one thing, Paterson and Passaic, in Passaic County, have large Black populations, but apparently not many majority tracts.) More damning is that nearly all majority-home-owner districts are majority-white.

```{r}
ggplot(filter(NNJ_tracts,pct_rental<0.5),aes(x = reorder(county, -1*pct_rental), y = pct_rental, color = majority_race_eth)) +
  geom_jitter(width = 0.35) +
  scale_x_discrete(name=NULL) +
  scale_y_continuous(name="Share of Rental Households", limits = c(0,1)) +
  scale_color_manual(name = "Majority Race/Ethnicity",
                       labels = c("Hispanic", "Non-Hispanic Asian",
                       "Non-Hispanic Black", "Non-Hispanic White",
                       "No Majority"),
                     values = c("darkorange2","olivedrab3","green","white","hotpink")) +
  ggtitle("Majority Homeowner, Majority Non-White")
```

(I'm not sure if it's worth including the majority-rental tracts or not...)
